---
title: "Solving a word funnel programming challenge"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{funnel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(funnel)
library(testthat)
```


## Introduction

This package solves the two of reddit r/dailyprogrammer challenges
[word funnel 1](https://www.reddit.com/r/dailyprogrammer/comments/98ufvz/20180820_challenge_366_easy_word_funnel_1/)
and
[word funnel 2](https://www.reddit.com/r/dailyprogrammer/comments/99d24u/20180822_challenge_366_intermediate_word_funnel_2/). We will be using them as a pedagogical tool to
explore advanced R concepts. Concepts include:

- Functional programming in R
    - Recursion
    - Closures
    - Pipes (`|>`)
    - Higher order functions
        - Applying functions over collections (map-like operations)
        - Passing functions into functions and returning functions from functions
- The Rcpp package
    - Including Rcpp in a package
    - Identifying performance bottlenecks in R code and rewriting them in C++
    - Passing around pointers to C++ objects within the R session using
      [`Rcpp::Xptr()`](https://www.r-bloggers.com/2010/01/external-pointers-with-rcpp/)
    - Implicit conversion between R types and C++ primitive, standard library,
      and standard template library (STL) types
- Data structures and algorithms
    - Comparing the performance of various search algorithms, including linear,
      binary, set membership, and hashmap membership.
    - Using the C++ standard library to gain access to advanced container types
      and algorithms not available in base R
    - Using non-standard R data types
      ([`hashtab()`](https://stat.ethz.ch/R-manual/R-devel/library/utils/html/hashtab.html))
      to improve performance
- Literate programming in R
    - Package-oriented [project compendium](https://www.jstor.org/stable/27594227)
    - Creating package datasets and tests to augment the compendium
    - Documenting supporting functions using
      [Roxygen](https://cran.r-project.org/package=roxygen2)
    - Vignettes as project write-ups using
      [Rmarkdown](https://cran.r-project.org/package=rmarkdown) (this document)

These toy problems will provide just enough complexity to warrant a
package-oriented project compendium and showcase its strengths, while remaining
simple enough so as not to get lost in project details. The challenges are also
easy to understand but of sufficient depth that they can be implemented a
naive way and a more sophisticated way to compare performance.

## Part 1

> Given two strings of letters, determine whether the second can be made from
the first by removing one letter. The remaining letters must stay in the same
order.

### Examples

```
funnel("leave", "eave") => true
funnel("reset", "rest") => true
funnel("dragoon", "dragon") => true
funnel("eave", "leave") => false
funnel("sleet", "lets") => false
funnel("skiff", "ski") => false
```

### Solution

In the trivial case, we can be assured the result is `FALSE` if the number of
characters in the second string is not one less than the first. Otherwise, we
will need to examine each string by indexing into it. In R, we can index into
a single string in a character vector, but we cannot index into a single
character into that string. So the approach here will be to split each string
using `strplit()`. This will likely cause a fairly significant performance hit
over a language like Python or C++ because it will result in memory allocations
for two new character vectors (plus a list to contain them). Not to mention the
overhead for doing the split. This challenge can be fairly simply completed in
R but it will not be efficient.

```r
r_funnel <- function(a, b) {
  # If nchar is not n - 1 then it is definitely wrong
  if (nchar(a) != nchar(b) + 1) return(FALSE)

  # Fragment each word into character vector
  strings <- strsplit(c(a, b), character(0))

  # Remove each character one by one and find out the resulting vector matches b
  any(
    vapply(
      seq_along(strings[[1]]),
      \(i) all(strings[[1]][-i] == strings[[2]]),
      logical(1L)
    )
  )
}
```

The Rcpp implementation is radically simpler. Note that Rcpp will automatically
convert a single-element R character vector into the C++ standard library's
string type `std::string`. From there, we can access each element of each string
using the `[]` operator to do our comparisons.

```c++
bool cpp_funnel(const std::string &a, const std::string &b) {
  size_t b_len = b.size();
  if (a.size() != b_len + 1) return false;

  bool skipped = false;

  for(size_t i=0, j=0; j < b_len;) {
    if (a[i] != b[j]) {
      if (skipped) return false;
      skipped = true;
      i++;
    }
    else i++, j++;
  }

  return true;
}
```

### Tests

This is the first instance where we demonstrate an advantage of using a
package-oriented project compendia. We can include automated tests to ensure
that our functions are performing as expected and run those tests with the
touch of a button. Assuming our project is already set up as a package, running
`usethis::use_test("part1")` will set up `testthat` in your project and create
a new test file for you to modify. We can put the following tests in the file
and then run <kbd>Ctrl</kbd> + <kbd>Shft</kbd> + <kbd>t</kbd> to run the tests.

```{r pt1-funnel-tests}
test_that("R funnel works", {
  r_funnel("leave", "eave")     |> expect_true()
  r_funnel("reset", "rest")     |> expect_true()
  r_funnel("dragoon", "dragon") |> expect_true()
  r_funnel("eave", "leave")     |> expect_false()
  r_funnel("sleet", "lets")     |> expect_false()
  r_funnel("skiff", "ski")      |> expect_false()
})

test_that("Rcpp funnel works", {
  cpp_funnel("leave", "eave")     |> expect_true()
  cpp_funnel("reset", "rest")     |> expect_true()
  cpp_funnel("dragoon", "dragon") |> expect_true()
  cpp_funnel("eave", "leave")     |> expect_false()
  cpp_funnel("sleet", "lets")     |> expect_false()
  cpp_funnel("skiff", "ski")      |> expect_false()
})
```

Yay! Both implementations return expected results for provided example data. 🥳

For more on creating packages, I recommend
[Hadley Wickam's _R Packages_](https://r-pkgs.org/) and the 
[`usethis` package](cran.r-project.org/package=usethis).

### Performance

It's good that we were able to solve this problem in pure R since there are
various reasons why we might not want to get Rcpp involved. For example,

- There is more overhead to get Rcpp set up.
- The compilation step requires build tools for the target machine, which are
  not always available by default
- C++ code is "lower level", which can make it more difficult to reason about
  a problem vs a "higher level" language like R which is designed for being
  expressive in a specific domain.
- R is portable and very shareable.

But there are times when either the advantages outweigh the disadvantages or
the disadvantages are minimized due to external factors. R does lots of tricks
under the hood to get good performance, so going to C++ is not always a silver
bullet. But some places where R might need some help are in string handling,
recursion and loops (sometimes), and data structures and algorithms.

The disadvantage of needing build tools is minimized when shipping code on
[CRAN]() since they will compile code to any target that may not have build
tools pre-installed. Packaging code itself is a strategy for making code more
shareable.

This challenge is a string handling problem, which was selected especially
because it can be one of R's weaker performance domains. We might expect a
significant difference in performance when benchmarking, and indeed we see a ~16
fold speed gain using C++. This could be very significant if this function is
expected to be used frequently, such as in a package meant for wide distribution
or if it needs to be used inside a tight loop (for example as part of an
exhaustive search algorithm).

```{r pt1-funnel-benchmark}
bench::mark(
  r_funnel("sleet", "lets"),
  cpp_funnel("sleet", "lets")
)

bench::mark(
  r_funnel("dragoon", "dragon"),
  cpp_funnel("dragoon", "dragon")
)
```

[enable1 word set](https://raw.githubusercontent.com/dolph/dictionary/master/enable1.txt)
